Spatial Reasoning and Data Displays

### 2015-06-23 Paper/InfoVis2015Revision/TestingVisualAptitude.Rnw
Changes during meeting

@@

Here, we want to investigate the extent to which observers make use of spatial ability in their evaluation of and reasoning about statistical graphics. As a quantitative measure of how successful an evaluation is, we are making use of the lineup protocol ~~\cite{buja2009statistical, wickham2010graphical, majumder2013validation}. 
\comment{Not sure where to add the Related Work section; the introduction is really the most similar thing, but adding another sub-section in would really kill the flow of the paragraph, in my opinion.}
\hhnote{I agree, it'll disrupt the flow, but we'll have to add a related work section nevertheless. We could add a bit of discussion on separating lineups from plots judging significance.}

\hhnote{ these are the references we are supposed to discuss (not sure, how to add the talbot one - that is a study extending cleveland and then heer's). first attempt: It has been shown~\cite{pak2014} that visualizations do not necessarily reflect statistical properties well or are even misleading viewers~\cite{correll2013}. Lineup tests are different as they do not ask participants to judge significances directly but are only based on comparisons. Statistical properties are then derived from this assessment.  This has been shown to be similarly powerful to standard statistical tests in situations where these classical tests hold, and more powerful in situations where conditions of classical tests are not met~\cite{majumder2013validation}. The difference between statistical properties and visualizations is very easy to explain, as statistical significance of, e.g. a correlation coefficient is based on only a few summary statistics, whereas e.g. a corresponding  scatterplot is showing a lot more details of the data which might lead participants to a different evaluation. This, however, is not a drawback of the visualization but only shows that summary statistics might not be enough to come to an appropriate conclusion \cite{anscombe:1973}. }
``Lineups" were introduced as a tool to evaluate the statistical significance of a graphical finding~\cite{wickham2010graphical}. Lineups are also useful in assessing the effectiveness of different graphical displays~\cite{hofmann2012graphical, loy:2015}. Like their police counterpart, lineups consist of several distractor plots (of randomly generated data) and one target (the data plot). 
% 
% \begin{figure}[htp]
% 
% \end{figure}

@@

In this paper, we present the results of a study designed to compare lineup performance with visual aptitude and reasoning tests, examining the skills necessary to successfully evaluate lineups. 
\newtext{
Our goal in this study is not to evaluate the statistical significance of particular plots, but rather, to consider the cognitive demands that we require of participants in lineup studies. 
}
We compare lineup performace to the visual search task (VST), paper folding test, card rotation test, and figure classification test. 
The VST measures visual search speed~\cite{goldstein1973validity}, the paper folding and card rotation tests measures spatial manipulation ability, and the figure classification test measures inductive reasoning~\cite{ekstrom1976manual}; 
all of these skills are at least peripherally recruited during the lineup task, but some may dominate in predicting performance on the lineup task. 
We hope to facilitate comparison of the lineup task to known cognitive tests, inform the design of future studies, and better understand the perception of statistical lineups.

@@

@@

The \textbf{paper folding test} measures participants' ability to visualize and mentally manipulate figures in three dimensions. A sample question from the test is shown in figure~\ref{fig:paperfolding}. It is classified as part of the visualization factor in~~\cite{ekstrom1976manual}, which differs from the spatial orientation factor because it requires participants to visualize, manipulate, and transform the figure~mentally, which makes it a more complex and demanding task than simple rotation. The paper folding test is associated with the ability to extrapolate symmetry and reflection over multiple steps.

@@

\subsection{Comparison of Spatial Tests with Previously Validated Results}
The card rotation, paper folding, and figure classification tests have been validated using different populations, many of which are demographically similar to Iowa State students (naval recruits, college students, late high-school students, and 9th grade students). We compare Iowa State students' unscaled scores in Table~\ref{tab:scorecomparison}, adjusting data from other populations to account for subpopulation structure and test length.

@@

<<VisReasoningCategorical, echo=F, include=F, eval=T, fig.width=8.75,fig.height=6, message=F, warning=F, fig.pos='ht'>>=
loadfonts()
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="verbal_skills", value=1))
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="math_skills", value=1))
lineup.summary.categorical$variable <- factor(lineup.summary.categorical$variable, labels=c("Age", "Sex", "Math/Science Research", "Statistics Class", "Calculus 1", "Verbal Skills", "Math Skills", "Art Skills", "AutoCAD Experience", "STEM Major", "Video Game Hrs/Wk"))
lineup.summary.categorical$value[lineup.summary.categorical$value=="f"] <- "Female"
lineup.summary.categorical$value[lineup.summary.categorical$value=="m"] <- "Male"
lineup.summary.categorical$value[lineup.summary.categorical$value=="y"] <- "TRUE"
lineup.summary.categorical$value[lineup.summary.categorical$value=="n"] <- "FALSE"

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="blank") +
  geom_boxplot(fill="grey", outlier.size=1.5) + 
  facet_wrap(~variable, scales="free_x", ncol=4) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=2.5) + 
  theme_bw() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="mm"))
if (file.exists("./Figure/fig-VisReasoningCategorical-1.pdf"))
  embedFonts("./Figure/fig-VisReasoningCategorical-1.pdf")
@
\begin{figure*}[h!tb]\centering
\includegraphics[width=.85\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}
%\afterpage{\clearpage}

@@

\subsection{Understanding Visual Abilities and Lineup Performance}
<<correlations, echo=F, comment="|", warning=F, message=F, results='asis', size="footnotesize", include=F>>=
print(xtable(cor(ans.summary[,c("lineup","card_rot", "fig_class", "folding", "vis_search")]), digits=c(1, rep(2, 5)), caption="Correlation matrix for the five tests.\\label{tab:corrmatrix5}"), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, size="footnotesize", scalebox=.9, table.placement="htb", caption.placement="top")
@

@@

<<pcarotation4,echo=F, results='asis'>>=
print(xtable(pca2$rotation, caption="Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\\label{tab:PCArotation4}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, table.placement="htb", caption.placement="top")
@

>>>>

% \hhnote{XXX Still missing: we need to extend the discussion a little bit that we want to extract the common core from lineups rather than investigate the relationship between individual designs and the  tasks. I believe, that the lineup commonalities are stronger than the design differences - but we might need to get some statistical backup for that. ... one way to do that would be to do a PCA on the three lineups and investigate whether anything beyond the first dimension is significant. I don't think so.  }

% \hhnote{XXX we should add: the lineups come from three different lineup experiments that discuss and compare the design choices (cite papers). This is NOT what we are doing here. Regardless of design choices, we want to find out, what visual skills are necessary for dealing with lineups. }

@@

Here, we want to investigate the extent to which observers make use of spatial ability in their evaluation of and reasoning about statistical graphics. As a quantitative measure of how successful an evaluation is, we are making use of the lineup protocol ~~\cite{buja2009statistical, wickham2010graphical, majumder2013validation}.

``Lineups" were introduced as a tool to evaluate the statistical significance of a graphical finding~\cite{wickham2010graphical}. Lineups are also useful in assessing the effectiveness of different graphical displays~\cite{hofmann2012graphical, loy:2015}. Like their police counterpart, lineups consist of several distractor plots (of randomly generated data) and one target (the data plot). 
% 
% \begin{figure}[htp]
% 
% \end{figure}

@@

In this paper, we present the results of a study designed to compare lineup performance with visual aptitude and reasoning tests, examining the skills necessary to successfully evaluate lineups. 
\newtext{
Our goal in this study is not to evaluate the statistical significance of particular plots or visualization designs, but rather, to consider the cognitive demands that we require of participants in lineup studies. 
}
We compare lineup performace to the visual search task (VST), paper folding test, card rotation test, and figure classification test. 
The VST measures visual search speed~\cite{goldstein1973validity}, the paper folding and card rotation tests measures spatial manipulation ability, and the figure classification test measures inductive reasoning~\cite{ekstrom1976manual}; 
all of these skills are at least peripherally recruited during the lineup task, but some may dominate in predicting performance on the lineup task. 
\newtext{
The cognitive tests from \cite{ekstrom1976manual} are highly validated tests that have been constructed to measure specific cognitive factors. We have additionally added the VST in order to additionally account for the basic visual search component of the lineup task.
}
We hope to facilitate comparison of the lineup task to known cognitive tests, inform the design of future studies, and better understand the perception of statistical lineups.

@@

\paragraph{Related Work}
\newtext{It has been shown~\cite{pak2014} that visualizations do not necessarily reflect statistical properties well or are even misleading viewers~\cite{correll2013}. Lineup tests are different as they do not ask participants to judge significances directly but are only based on comparisons. Statistical properties are then derived from this assessment.  This has been shown to be similarly powerful to standard statistical tests in situations where these classical tests hold, and more powerful in situations where conditions of classical tests are not met~\cite{majumder2013validation}. The difference between statistical properties and visualizations is very easy to explain, as statistical significance of, e.g. a correlation coefficient is based on only a few summary statistics, whereas e.g. a corresponding scatterplot is showing a lot more details of the data which might lead participants to a different evaluation. This, however, is not a drawback of the visualization but only shows that summary statistics might not be enough to come to an appropriate conclusion \cite{anscombe:1973}. In addition, the lineup task does not require participants to make numerical judgments, as in \cite{talbot}, but rather to indicate (though not identify) a visual difference; this leads to a much more natural testing procedure.}
% \hhnote{XXX still missing: talbot reference; not sure, how to add the talbot one - it is a really nice study  extending cleveland's and then heer's, but not directly related.}

@@

The \textbf{paper folding test} measures participants' ability to visualize and mentally manipulate figures in three dimensions. A sample question from the test is shown in figure~\ref{fig:paperfolding}. It is classified as part of the visualization factor in~~\cite{ekstrom1976manual}, which differs from the spatial orientation factor because it requires participants to visualize, manipulate, and transform the figure~mentally, which makes it a more complex and demanding task than simple rotation. The paper folding test is associated with the ability to extrapolate symmetry and reflection.

@@

\subsection{Comparison of Spatial Tests with Validated Results}
The card rotation, paper folding, and figure classification tests have been validated using different populations, many of which are demographically similar to Iowa State students (naval recruits, college students, late high-school students, and 9th grade students). We compare Iowa State students' unscaled scores in Table~\ref{tab:scorecomparison}, adjusting data from other populations to account for subpopulation structure and test length.

@@

<<VisReasoningCategorical, echo=F, include=F, eval=T, fig.width=6,fig.height=5, message=F, warning=F, fig.pos='ht'>>=
loadfonts()
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="verbal_skills", value=1))
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="math_skills", value=1))
lineup.summary.categorical$variable <- factor(lineup.summary.categorical$variable, labels=c("Age", "Sex", "Math/Science Research", "Statistics Class", "Calculus 1", "Verbal Skills", "Math Skills", "Art Skills", "AutoCAD Experience", "STEM Major", "Video Game Hrs/Wk"))
lineup.summary.categorical$value[lineup.summary.categorical$value=="f"] <- "Female"
lineup.summary.categorical$value[lineup.summary.categorical$value=="m"] <- "Male"
lineup.summary.categorical$value[lineup.summary.categorical$value=="y"] <- "TRUE"
lineup.summary.categorical$value[lineup.summary.categorical$value=="n"] <- "FALSE"

lineup.summary.categorical$variable <- gsub("Math/Science Research", "Math/Sci Research", lineup.summary.categorical$variable)

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="blank") +
  geom_boxplot(fill="grey", outlier.size=1.5) + 
  facet_wrap(~variable, scales="free_x", ncol=4) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=2.5) + 
  theme_bw() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="mm"))
if (file.exists("./Figure/fig-VisReasoningCategorical-1.pdf"))
  embedFonts("./Figure/fig-VisReasoningCategorical-1.pdf")
@
\begin{figure}[h!tb]\centering
\includegraphics[width=.95\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure}
%\afterpage{\clearpage}

@@

\subsection{Visual Abilities and Lineup Performance}
<<correlations, echo=F, comment="|", warning=F, message=F, results='asis', size="footnotesize", include=F>>=
print(xtable(cor(ans.summary[,c("lineup","card_rot", "fig_class", "folding", "vis_search")]), digits=c(1, rep(2, 5)), caption="Correlation matrix for the five tests.\\label{tab:corrmatrix5}"), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, size="footnotesize", scalebox=.9, table.placement="htb", caption.placement="top")
@

@@

<<pcarotation4,echo=F, results='asis'>>=
row.names(pca2$rotation) <- c("Card Rot", "Fig Class", "Folding", "VST")
print(xtable(pca2$rotation, caption="Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\\label{tab:PCArotation4}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, table.placement="htb", caption.placement="top")
@
